Mechanized Reasoning about a Capability Machine

### 2019-10-18 prisc-2020/paper.tex
Nitpicks

In summary, we present Iris formalizations of 
\begin{itemize}
  \item a program logic for reasoning about capability machine programs.
  \item the logical relation from \cite{skorstengaardESOP18}, 
    which captures capability safety and which can be used to reason about examples
    that rely on WBCF and LSE.
\end{itemize}
Most of the technical development is mechanized in Coq using the Coq implementation of Iris (the remaining work consists of a few missing details). Currently the mechanization consists of XXX lines of code. Its substantial size can be slightly improved by some clever use of tactics, however such a mechanization will always be large, due to the low level nature of a machine: we want to consider every possible case of an instruction, both the fails and the successes.

@@

Next we use the Iris framework to \emph{prove} a Hoare triple for each instruction.
As in \cite{Jensen:2013}, we define a points-to predicate for registers, denoted $r \mapsto_r w$.
Since a capability machine replaces pointers with capabilities, we replace the conventional points-to predicate of separation logic for pointers with a points-to predicate with a permission attached to it, denoted
$a \mapsto_a[p]~w$. 
% \[
%    \hspace{1cm} a \mapsto_a[p]~w.
% \]
This predicate states that address $a$ points to word $w$ with permission $p$.
The permission restricts how the memory may be updated at address $a$.
For instance, if $a \mapsto_a[\textsc{\tiny RX}]~w$, then the $\textsc{\tiny RX}$ (ReadExecute)
permission gives us permission to read from address $a$, to execute w, but it does not to permit us to write to address $a$.

% Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
The Hoare triples for basic instructions take the following form
\begin{align*}
\small
	\textit{run time conditions } \wedge &\textit{ decode}(w) = \emph{instr} \Rightarrow  \\
	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
	\textsf{Instr Executable }\\
	\{\{\{\textsf{PC} \mapsto_r ((p,g),b,e,a&+1) * a \mapsto_a[p] w * ... \}\}\}
\end{align*} 
Here, the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
This form of Hoare triple is similar to the one used in \cite{Jensen:2013} but
unlike \cite{Jensen:2013}, the decoding function is in our case assumed.

@@

In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
We use these relations in combination with Iris' saved predicates, which allow us to save a predicate by associating it to a unique name.
Using that unique name, we are then able to refer to the saved predicate somewhere else, and apply it to
an appropriate future world argument.

>>>>

In summary, we present Iris formalizations of 
\begin{itemize}
  \item a program logic for reasoning about capability machine programs.
  \item the logical relation from \cite{skorstengaardESOP18}, 
    which captures capability safety and which can be used to reason about examples
    that rely on WBCF and LSE.
\end{itemize}
Most of the technical development is mechanized in Coq using the Coq implementation of Iris (the remaining work consists of a few missing details). Currently the mechanization consists of 5.8k lines of spec and 15k lines of proof as reported by \textsf{coqwc}. Its substantial size can be slightly improved by some clever use of tactics, however such a mechanization will always be large, due to the low level nature of a machine: we want to consider every possible case of an instruction, both the fails and the successes.
\todo{Alix: I would say it's more due to the nature of a capability machine with dynamic checks and the multiple ways they can fail}

@@

Next we use the Iris framework to \emph{prove} Hoare triples for each instruction.
As in \cite{Jensen:2013}, we define a points-to predicate for registers, denoted $r \mapsto_r w$.
Since a capability machine replaces pointers with capabilities, we replace the conventional points-to predicate of separation logic for pointers with a points-to predicate with a permission attached to it, denoted
$a \mapsto_a[p]~w$. 
% \[
%    \hspace{1cm} a \mapsto_a[p]~w.
% \]
This predicate states that address $a$ points to word $w$ with permission $p$.
The permission restricts how the memory may be updated at address $a$.
For instance, if $a \mapsto_a[\textsc{\tiny RX}]~w$, then the $\textsc{\tiny RX}$ (ReadExecute)
permission gives us permission to read from address $a$, to execute w, but it does not to permit us to write to address $a$.

% Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
The Hoare triples for basic instructions take the following form
\begin{align*}
\small
	\textit{run time conditions } \wedge &\textit{ decode}(w) = \emph{instr} \Rightarrow  \\
	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
	\textsf{Instr Executable }\\
	\{\{\{\textsf{PC} \mapsto_r ((p,g),b,e,a&+1) * a \mapsto_a[p] w * ... \}\}\}
\end{align*} 
Here, the runtime conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
This form of Hoare triple is similar to the one used in \cite{Jensen:2013} but
unlike \cite{Jensen:2013}, the decoding function is in our case assumed.

@@

In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
Skorstengaard et al.~\cite{skorstengaardESOP18} distinguish between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
We use these relations in combination with Iris' saved predicates, which allow us to save a predicate by associating it to a unique name.
Using that unique name, we are then able to refer to the saved predicate somewhere else, and apply it to
an appropriate future world argument.
